---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

### Isaac Dorfman
### September 18, 2017

## Loading and preprocessing the data

To insure maximum portability of code the importing and preprocessing of the  
data handled as such:

```{r}
tmp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",tmp)
activity <- read.csv(unz(tmp, "activity.csv"))
```

In the event that the URL for the data set changes in the future this portion  
of the code is easily updated for such contingency.  In the event that a  
future user of this code already has the data set stored locally then the  
read.csv function can be used by itself.

## What is mean total number of steps taken per day?

At this step in the analysis it is permissable to ignore the copious numer of  
NAs in the data set in order to get a better idea of what information we  
actually have.

```{r}
activity_mean <- mean(activity$steps, na.rm = TRUE)
activity_mean

activity_median <- median(activity$steps, na.rm = TRUE)
activity_median
```

We can also examine a histogram of the data to see why the mean is likely  
going to be more informative of further analysis than the median:

```{r plot1, eval = TRUE, include = TRUE,fig.path = "figure/",fig.keep = "high", fig.show = "asis"}
hist(activity$steps, col = "steelblue", xlab = "Steps Taken Per Day",
     main = "Histogram of the Number of Steps Taken Per Day")

```

From this we can see that for the days for which we have data that the  
overwhelming majority of days saw no activity from the device.

## What is the average daily activity pattern?

To get a better impression of the daily pattern the first step is going to be  
a simple (and somewhat messy) time series plot:

```{r plot2, eval = TRUE, include = TRUE,fig.path = "figure/",fig.keep = "high", fig.show = "asis"}
plot(activity$interval,activity$steps, type = "l", xlab = "Interval", 
     ylab = "Number of Steps", main = "Count of Steps by Interval")
```

We would also like to know which interval saw the maximum number of steps:

```{r}
activity_max <- activity[which(activity[,"steps"]==max(activity$steps, na.rm = TRUE)),]
activity_max
```

There are many different approaches to finding both the maximum value for a  
given variable and the associated observation.  This approach was selected  
because it would allow us to see if the maximum value occurred more than once.  
While it, as can be seen above, does not it is still a good idea to check for  
such possibilities.

## Imputing missing values

Before we can begin any imputation for missing values it would be helpful to  
know how many missing values we are dealing with and what percent of the total  
data set they represent.

```{r}
activity_missing <- sum(!complete.cases(activity))
activity_missing
totals_obs <- nrow(activity)
percent_missing <- 100*(activity_missing/totals_obs)
percent_missing
```

It looks like a little over 13% of our data set is missing a value for the  
'steps' variable.  The next step is to come up with a reasonable estimate to  
fill in those missing values.  After some research on this topic the 'mice'  
package was selected to assist with creating an imputed data set to fill in the  
missing values.

```{r, echo=FALSE}
packages <- c("lattice", "mice")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}
```

```{r}
require("mice")

md.pattern(activity)
```

This verifies our earlier calculations for the number of missing values.  Time  
to impute their fill ins.  

```{r}
activity.imp <- mice(activity, m=1, method = "pmm", seed = 11235)
activity_complete <- complete(activity.imp,1)
mean(activity_complete$steps)
median(activity_complete$steps)
```

```{r plot3, eval = TRUE, include = TRUE,fig.path = "figure/",fig.keep = "high", fig.show = "asis"}
hist(activity_complete$steps, col = "paleturquoise1", xlab = "Steps Taken Per Day 
     (with imputed values included)",
     main = "Histogram of the Number of Steps Taken Per Day")
```

Even with the imputed values substituting for NAs the median is still 0 while  
the imputed mean is around three steps per five minute interval higher than  
the original mean.

## Are there differences in activity patterns between weekdays and weekends?

Since we are looking at activity levels it would be helpful to know if there  
are predictible periods of higher or lower activity than normal or if there is  
a predictible split in activity levels.  To this end a comparison of weekdays  
against the weekend.

```{r}
activity_complete$date <- strptime(activity_complete$date, "%Y-%m-%d")
days <- c("Monday","Tuesday","Wednesday","Thursday", "Friday")
activity_complete$weekday <- factor((weekdays(activity_complete$date) %in% days), 
                           levels=c(FALSE, TRUE), 
                           labels=c('weekend', 'weekday'))
```

Now that we have the factor variables 'weekday' we can creat a plot to see if  
there are any differences between weekdays and weekends.

```{r plot4, eval = TRUE, include = TRUE,fig.path = "figure/",fig.keep = "high", fig.show = "asis"}
require("lattice")

xyplot(steps~interval | weekday, data = activity_complete, type="l", 
       col="lightcyan3", layout = c(1,2))
```
